### 2025-08-15 basic-statistics.qmd
20250815 - correlation updates

$$
\begin{aligned}
  r &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{the numerator is the covariance; the denominator is the standard deviations} \\
    &= \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{$n-1$ cancels out in numerator and denominator (typical formula for Pearson correlation)} \\
    &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{s_x \cdot s_y} 
  && \quad \text{adding back $n-1$, replacing denominator from line 1 with standard deviation symbols} \\
    &= \frac{\text{Cov}(x, y)}{s_x \cdot s_y} 
  && \quad \text{replacing numerator with covariance notation}
\end{aligned}
$$ {#eq-pearsonCorrelation}

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $\text{Cov}(x, y)$ is the covariance between $x$ and $y$, $s_x$ is the standard deviation of $x$, and $s_y$ is the standard deviation of $y$.

>>>>

$$
\begin{aligned}
  r &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\frac{1}{n - 1} \sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{the numerator is the covariance; the denominator is the standard deviations} \\
    &= \frac{\sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}
{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2} \sqrt{\sum_{i=1}^{n} (y_i - \bar{y})^2}} 
  && \quad \text{$n-1$ cancels out in numerator and denominator (typical formula for Pearson correlation)} \\
    &= \frac{\frac{1}{n - 1} \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{s_x \cdot s_y} 
  && \quad \text{adding back $n-1$, replacing denominator from line 1 with standard deviation symbols} \\
    &= \frac{\text{Cov}(x, y)}{s_x \cdot s_y} 
  && \quad \text{replacing numerator with covariance notation} \\
    &= \frac{\sum z_x z_y}{n - 1}
\end{aligned}
$$ {#eq-pearsonCorrelation}

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $\text{Cov}(x, y)$ is the covariance between $x$ and $y$, $s_x$ is the standard deviation of $x$, $s_y$ is the standard deviation of $y$, $n$ is the sample size, $z_x$ is the *z*-score for $x$, and $z_y$ is the *z*-score for $y$.

### 2025-08-15 correlation.qmd
20250815 - correlation updates

Also note that, although the ([Pearson](#sec-correlationExamplesPearson)) correlation reflects the strength of the (linear) association between two variables, it does not reflect the slope of that association, as depicted in @fig-correlationExamples.

@@

@@

@@

The correlation coefficient is strongly impacted by outliers—i.e., extreme (i.e., very large or very small) values relative to the other observations.
Consider the following example:

>>>>

Also note that, although the ([Pearson](#sec-correlationExamplesPearson)) correlation reflects the strength of the (linear) association between two variables, it does not reflect the (unstandardized) slope of that association, as depicted in @fig-correlationExamples.

@@

The [Pearson](#sec-correlationExamplesPearson) correlation coefficient can be greatly impacted by the extent of variability in the data, differences in the shapes of the two distributions, [nonlinearity](#sec-correlationExamplesNonlinear), [outliers](#sec-correlationOutliers), characteristics of the sample, and measurement error [@Goodwin2006].
These factors tend to have greater impact when the sample size is smaller (compared to when the sample size is larger).
The [Pearson](#sec-correlationExamplesPearson) correlation coefficient tends to be smaller when:

- there is low variability in one or both variables (i.e., [restricted range](#sec-correlationRestrictedRange))
- the two distributions show dissimilar shapes (in terms of [skewness and kurtosis](#sec-descriptiveStatisticsShape))
- the underlying association is nonlinear, such that [Pearson](#sec-correlationExamplesPearson) correlation does not capture the association
- [outliers](#sec-correlationOutliers) weaken the apparent linear trend
- the association differs across subgroups of the sample
- the data contain random measurement error

By contrast, the association between two variables can be artificially inflated by factors such as:

- [outliers](#sec-correlationOutliers) that happen to fall along a linear trend
- systematic measurement error (e.g., common method bias)
- third variable [confounding](#sec-causalDiagramConfounding)

@@

You can compute a correlation manually using the *z*-scores, as in @eq-pearsonCorrelation:

```{r}
sum(scale(complete_cases$height) * scale(complete_cases$fantasyPoints)) / nrow(complete_cases)
```

@@

The correlation coefficient is strongly impacted by outliers—i.e., extreme (i.e., very large or very small) values relative to the other observations.
Outliers can arise from errors that arise from data collection, measurement, or data entry, or they could represent valid (albeit atypical) values [@Goodwin2006].
Outliers can either artificially weaken the estimate of association—if they weaken the apparent linear trend—or artificially strengthen it—if they happen to fall along a linear trend.
Consider the following example:
